# A flexible programming language

::: {.goals}
-   The core data types are numeric, logical, and character
-   The most important data structures are vectors and data frames
-   Vectors store a multitude of the same data type in a particular order
-   Data frames are tables with data organized in rows and columns
-   To assign data to variables use `<-`
-   Write comments with `#`
:::

R is a powerful and complex programming language that supports many different programming
paradigms. However, to get started with data analysis, its enough to understand R's basic
data types (numeric, logical, and character) and two data structures: vectors and data
frames. These are made for data manipulation and analysis, and along with the many
convenience functions that R provides, they allow you to focus on data analysis without
worrying about thing that are fundamental in other languages like [control
flow](https://en.wikipedia.org/wiki/Control_flow) (loops and if-else-statements). As you
become more familiar with R, you can add those to your repertoire, too, and utilize R's
full power and flexibility.

## Data Types

The [data type](https://en.wikipedia.org/wiki/Data_type) of a piece of data determines
what operations you can perform with it. The bread and butter of data types in R are
*numeric*, *logical*, and *character*.

#### Numeric {.unnumbered}

Numeric data can be integer, decimal[^flexible-1], or complex, but usually the difference
does not matter. Sometimes, R might unexpectedly hit you with a number in scientific
notation, e.g., `1e+23`. To the uninitiated, these can look cryptic and a bit scary. But
don't worry! It's just a very large or very small number that is written in a more compact
form and you can [change disable this if you
like!](https://statisticsglobe.com/disable-exponential-scientific-notation-in-r)

[^flexible-1]: Decimal numbers are also called floating point numbers or "float" for
    short. Note that R uses periods `.` to designate the decimal places, not commas `,` as
    in some countries.

```{r, eval=FALSE}
23 + 0.45 + 1e2 + 6i
```

#### Logical {.unnumbered}

Logical data (also known as Boolean data) can be either `TRUE` or `FALSE` (or their
shorthands, `T` and `F`). These are often used to represent binary categorical variables,
as function arguments, or as intermediate steps when selecting and filtering data.

```{r, eval=FALSE}
(TRUE || FALSE) != (F == T)
```

#### Character {.unnumbered}

Also known as *strings*. Almost anything you type becomes a character string if you
surround it with `""` or `''`.

```{r, eval=FALSE}
"Hello World!"
```

## Assignment

To assign values to *variables*[^flexible-2], you can use `<-`. Use the keyboard shortcut
<kbd>Alt</kbd>+<kbd>-</kbd> to save yourself some hand gymnastics. Note how the newly
created variables appear in the environment panel in RStudio. Once assigned, you can use
variable's name to represent its content in other operations. Typing a variable name by
itself will print the variable's content.

[^flexible-2]: In official R lingo you are assigning values to *objects*, but people who
    have experience with other programming languages sometimes find this term confusing.
    Therefore, I will use the word *variable* in this book.

```{r}
a <- 1
b <- 9
c <- a + b

c
```

::: {.more}
Many other programming languages use `=` to assign values to variables. You can, for the
most part, also use `=` for assignment in R --- just be aware that it will let everyone
now that [you are not a true believer](https://en.wikipedia.org/wiki/Shibboleth).

You can also use `->` to assign from left to right. This can it come in handy when you are
messing around in the console, but it may not be a good idea to use it too much in your
scrips, because it can make your code harder to read:

```{r eval=FALSE}
"Never look back!" -> x
```

Finally, there's `<<-`, which allows assignment to variables in an enclosing environment.
This can be useful if you need to update *global variables* from within a function.
However, this is far beyond the scope of this intro, and chances are you will never
encounter a `<<-` in the wild.

You can read more about R's assignment operators
[here](https://colinfay.me/r-assignment/).
:::

There are some rules for variables names: They must start with a letter and contain only
the letters of the English alphabet, numbers, underscores `_`, hyphens `-`, and dots
`.`.[^flexible-3] For long variable names, [it is recommended to use "snake
case"](https://style.tidyverse.org/syntax.html#object-names), which looks like this:
`my_firssst_variable` ðŸ. Also note that R is case-sensitive.

[^flexible-3]: Unlike in other programming languages, dots have no special function or
    meaning and can be used in variable names.

::: {.exercise}
What would be the result of running the code below? Think about it first and then run it
to check your answer.

```{r eval=FALSE}
a <- 10
a + b    # What will be the result of a + b?
c        # What has happened to c?
C        # Why is this not working?
```
:::

## Comments

As you can see in the example above, you can use `#` to leave comments in your script.
[Learn to write good comments](https://www.youtube.com/watch?v=yhF7OmuIILc) and write
comments gratuitously -- your collaborators and future self will thank you for it!

## Vectors

In computer programming, [**data
structures**](https://en.wikipedia.org/wiki/Data_structure) are a way to organize and
process particular types of data efficiently. When working with data in R, you will
encounter two important data structures: *vectors* and *data frames*. Vectors are made for
storing values *of the same data type* (numerical, character, logical, ...) in a
particular order.

To create a vector, you can use the "combine" function, `c()`. There are also the `:` and
`seq()` functions, that can help you create a vector that contains a sequence of numbers:

```{r, results="hold"}
c(1,2,3,4)      # create a vector manually
0:4             # a vector of all integer numbers from 0 to 4
4:0             # it also works backwards
seq(0, 2, 0.5)  # numbers 1.0 to 2.0 in 0.5 increments
```

R is really good at working with vectors! See how easy it is to multiply all numbers by a
scalar, get the mean, or sum all elements --- all without using a loop or a temporary
variable:

```{r, results = "hold"}
x <- c(1,2,3,4) # Create vector and assign to variable `x` 
x * 10
mean(x) 
sum(x)
```

You can access specific positions of a vector using square brackets. Use `:` within
brackets to access a range of indices. In R, indices start at 1, that is, index 1 will
return the first element of the vector.

```{r}
x[1]

x[2:3]
```

You can also use a vector of indices to index into another vector in order to select
specific elements

```{r}
x[c(1,3)]   
```

You can also use the bracket syntax to replace individual elements:

```{r}
x[2:3] <- 0   # replace 2nd and 3rd element with 0
x[2:3] <- c(7,8)   # replace 2nd and 3rd element with 7 and 8, respectively
```

Values in a vector can be named. This comes in handy, for example, if you need a simple
[key-value store:](https://en.wikipedia.org/wiki/Key%E2%80%93value_database)

```{r}
fav_food <- c(
    Garfield = "Lasagna", 
    Popeye = "Spinach",
    Obelix = "Roast Wild Boar"
)

fav_food["Popeye"]
```

::: {.more}
Vectors can only contain a *single* data type. If you are trying to mix data types, R will
quietly (without throwing an error) *coerce* all data types to a data type that can
represent all elements, following a certain hierarchy of coercion:

-   logical becomes numerical (0 and 1)
-   numerical becomes character

This coercion only goes in one direction. For example, character strings that happen to
contain numbers will nevertheless be treated as characters. If you want R to treat a
character representation of a number as numeric, you have to explicitly tell it to do so,
for example with the `as.numeric()` function.

Consider these examples:

```{r}
c(1, 2, "3")
c(FALSE, TRUE, 2, 3)

x <- 1:3 
x[1] <- TRUE 
x 
```

Vectors also cannot contain other vectors, so nested vectors automatically get unnested:

```{r}
c( c(1,2), c(3,4) ) 
c( 1, c(2,3,4) ) 
c( c(TRUE, 0), c(TRUE, FALSE), c(TRUE, "0") )
```

If you need to store a sequence of mixed data types, you will have to use another data
structure called *list*.
:::

## Data Frames

Usually, your research data will be more complex than just a single sequence of values.
For this, R has data frames. Data frames are two-dimensional like a spreadsheet -- with
*rows* usually containing different observations and *columns* usually containing
different variables.

Rows and columns of a data frame can be named, and while row names seem to have fallen out
of fashion, it is good practice to give the columns of your data frames names that are
easy to remember and easy to type.

You can make your own data frame by hand, by putting together vectors like this:

```{r}
cool_numbers <- data.frame(
    "number" = c(1,13,23,42),
    "isPrime" = c(FALSE, TRUE, TRUE, FALSE)
)
```

A quick way to access a single column of a data frame is with the `$` operator.

```{r}
cool_numbers$number
```

Did you notice that columns in a data frame are vectors? This means that everything that
can be done with a vector can be done with a data frame column!

::: {.more}
In fact, all "atomic" data types in R (numeric, characters, logical) are *always* stored
as vectors -- even a single number is actually a numerical vector of length one! This is a
really powerful idea and the source of much of R's strength!
:::

::: {.exercise}
Make a new data frame `my_cool_numbers` with a column `a` that contains all numbers from 1
to 100 and column `b` that contains your favorite number 100 times.

::: {.answer}
``` {.r}
my_cool_numbers <- data.frame(
    "a" = 1:100,
    "b" = 42
)
```
:::

Since data frame columns are vectors, we can use the same operations on them. calculate
the standard deviation of all numbers in column `a` in our `my_cool_numbers` data frame!
Use the RStudio help to find the right function for it.

::: {.answer}
``` {.r}
sd(my_cool_numbers$a)
```
:::
:::

As for vectors, R comes out of the box with many more commands to create and modify data
frames, select columns, and filter rows. You will eventually need to get familiar with
these. However, in this short introduction we will skip these in favor of the more
streamlined and beginner-friendly functions from the Tidyverse -- more about that later.

## Other Data Structures

R has a couple of other data structures that you will eventually encounter. I don't want
to dwell on them for too long, because none of them will play a role in the rest this
book. If you are interested you can read more [here](http://adv-r.had.co.nz/). Additional
R data structures include ...

-   **Factor** -- Factors are numerical vectors that can take a fixed set of values
    (levels), whereby each level can be assigned a label. The levels may be ordered or
    unordered. If this reminds you of categorical and ordinal variables you are absolutely
    right -- that's where they are used most often. While factors can be quite useful,
    they are also a point of confusion for many R novices.
-   **List** -- Lists are similar to vectors, but they can contain a mix of data types and
    data structures, including other lists.
-   [**Tibble**](https://tibble.tidyverse.org/) -- A 'modern' implementation of data
    frames.
-   **Object** -- Objects can be used to organize functionality. For instance, a
    statistical model may be an object that has various attributes and associated
    functions. R even has three types of objects: R6, S3, and S4.
-   **Matrix** -- Not to be confused with data frames. If you are implementing statistical
    methods yourself, you may want to make use of R's matrix algebra functions. Otherwise,
    you likely won't often have to work with matrices directly.
-   [**Environment**](https://adv-r.hadley.nz/environments.html) -- An environment is
    similar to a names list, in that it contains a bunch of named R data structures
    Understanding environments is central to a deep understanding of R, but as a beginner
    it's nothing you need to worry about.
