# A flexible programming language that grows with you

You need very little [boilerplate code](https://en.wikipedia.org/wiki/Boilerplate_code) 
to set up a new R project and start working efficiently -- and sometimes the console is all you need.

R also grows with your needs: Beginners don't need to worry about many of the programming concepts
that are fundamental to working in other languages, such as [control
flow](https://en.wikipedia.org/wiki/Control_flow), because R's vector-based
approach to data manipulation will go a long way and many convenience functions
take care of the nitty-gritty for you. Then, as you become better and more cofortable with R, you can start to utilize its full power and flexibility. While R is a [functional
language](https://en.wikipedia.org/wiki/Functional_programming), it supports
different programming paradigms. It also allows powerful *meta-programming*
techniques that are much more difficult to achieve in other languages.

## Data Types

The [data type](https://en.wikipedia.org/wiki/Data_type) determines what
operations you can perform with a piece of data. The bread and butter of data
types in R are *numeric*, *logical*, and *character*.

### Numeric

Numbers can be integer, decimal, or complex, but usually the difference does not
matter. Sometimes, R might unexpectedly hit you with a number in scientific
notation, e.g., `1e+23` -- don't worry, it's still just a number and you can
[change the formatting if you
like!](https://statisticsglobe.com/disable-exponential-scientific-notation-in-r)

```{r, eval=FALSE} 
23 + 0.45 + 1e2 + 6i
```

### Logical

Can be either `TRUE` or `FALSE` (or their shorthands, `T` and `F`). These are
often used to represent binary categorical variables, as function arguments, or
as intermediate steps when selecting and filtering data.

```{r, eval=FALSE} 
(TRUE || FALSE) != (F == T)
```

### Character

Also known as "strings". Almost anything you type becomes a character string if
you surround it with `""` or `''`.

```{r, eval=FALSE} 
"Hello World!"
```

## Assignment

You can use `<-` to assign values to *variables* (*objects*, in R lingo). Use
the keyboard Shortcut <kbd>Alt</kbd>+<kbd>-</kbd> to save yourself some hand
gymnastics. Note how the newly created variables appear in the environment panel in RStudio.

You can give your variables almost any name you want as long as the
name starts with a letter. For long variable names, [it is recommended to use
"snake case"](https://style.tidyverse.org/syntax.html#object-names), which looks
like this: `my_firssst_variable` 🐍

```{r}
a <- 1
b <- 9
c <- a + b

c   # A variable name by itself will print the variable's content
```

:::{ .exercise }
What would be the result of running this piece of code now? Think about it first and then run it to check your answer.

```{r eval=FALSE}
a <- 10
a + b    # What is a + b now?
c        # What has happened to c?
```
:::

 
:::{ .digression } 
Many other programming languages use `=` to assign values to variables. In R,
you can (for the most part) also use `=` for assignment. Just be aware that it
will let everyone now that [you are not a true
believer](https://en.wikipedia.org/wiki/Shibboleth).

You can also use `->` to assign from left to right: `"I'm feeling freaky!" ->
x`. It may not be a good idea to use this in your scripts too much, because it
can make your code harder to read, but it sometimes comes in handy when you are
messing around in the console.

Finally, there's `<<-`, which allows assignment to variables in an enclosing
environment. This can be useful if you need to update *global variables* from
within a function. However, this is far beyond the scope of this intro, and
chances are you will never encounter a `<<-` in the wild.

You can [read more about R's assignment operators
here](https://colinfay.me/r-assignment/) 
:::

## Comments

As you can see in the example above, you can use `#` to leave comments in your
script. [Learn to write good comments](https://www.youtube.com/watch?v=yhF7OmuIILc) and write comments
gratuitously -- your collaborators and future self will thank you for it!

## Vectors

In computer programming, [**data structures**](https://en.wikipedia.org/wiki/Data_structure) 
are a way to organize and process particular types of data efficiently. 
When working with data in R, you will encounter two important data structures: *vectors* and 
*data frames*. Vectors are made for storing values *of the same data
type* (numerical, character, logical, ...) in a particular order.

To create a vector, you can use the "combine" function, `c()`. There are also
the `:` and `seq()` functions, that can help you create a vector that contains a
sequence of numbers:
    
```{r, results="hold"}
c(1,2,3,4)      # create a vector manually
0:4             # a vector of all integer numbers from 0 to 4
4:0             # it also works backwards
seq(0, 2, 0.5)  # numbers 1.0 to 2.0 in 0.5 increments
```

R is really good at working with vectors! See how easy it is to multiply all
numbers by a scalar, get the mean, or sum all elements --- all without using a
loop or a temporary variable:

```{r, results = "hold"} 
x <- c(1,2,3,4) # Create vector and assign to variable `x` 
x * 10
mean(x) 
sum(x)
```

You can access specific positions of a vector using square brackets. 
Use `:` within brackets to access a range of indices.  
In R, indices start at 1, that is, index 1 will give you the first element of the vector. 

```{r} 
x[1]
x[2:3]
x[c(1,3)]     # use a vector to select specific indices
```

You can also use the bracket syntax to replace individual elements:

```{r}
x[2:3] <- 0   # replace 2nd and 3rd element with 0
``` 

Values in a vector can be named. This comes in handy, for example, if you need a
simple [key-value store:](https://en.wikipedia.org/wiki/Key%E2%80%93value_database)

```{r}
fav_food <- c(
    Garfield = "Lasagna", 
    Popeye = "Spinach",
    Obelix = "Roast Wild Boar"
)

fav_food["Popeye"]
```


:::{ .more } 
*Here be dragons ...*

When working with vectors, there are some pitfalls. Consider these examples:

```{r}
c(1, 2, "a")
c(FALSE, TRUE, 2, 3)
```

```{r}
x <- 1:3 
x[1] <- TRUE 
x 

x[3] <- "3" 
x
```

```{r}
c( c(1,2), c(3,4) ) 
c( 1, c(2,3,4)) 
c( c(TRUE, 0), c(TRUE, FALSE), c(TRUE, "0"))
```

What is happening here? Remember that vectors can only contain a *single* data
type. If you are trying to mix data types, R will quietly (without throwing an
error) *coerce* all data types to the lowest common denominator: Logical become
numerical (0 and 1) and numerical become character. Vectors also can't contain
other vectors, so these automatically get unnested.

Character strings that happen to contain numbers will nevertheless be treated
as characters. If you want R to treat them as numeric, you have to
explicitly tell it to do so with the `as.numeric()` function.

If you need to store a sequence of mixed data types, you should not use a vector but 
a data structure called *list*. 
:::

## Data Frames

Usually, your research data will be more complex than just a single sequence of
values. For this, R has data frames. Data frames are two-dimensional like a
spreadsheet -- with *rows* usually containing different observations and
*columns* usually containing different variables.

Rows and columns of a data frame can be named, and while row names have fallen
out of fashion, it is good practice to give the columns of your data frames
names that are easy to remember and easy to type.

```{r}
# You can make your own data frame by hand like this
cool_numbers <- data.frame(
    "number" = c(1,13,23,42),
    "isPrime" = c(FALSE, TRUE, TRUE, FALSE)
)
```

A quick way to access a single column of a data frame is with the `$` operator

```{r}
cool_numbers$number
```

As for vectors, R comes out of the box with many more commands to create and
modify data frames, select columns, and filter rows. You will eventually need to
get familiar with these. However, in this short introduction we will skip these
in favor of the more streamlined and beginner-friendly functions from the
Tidyverse -- more about that later.

:::{ .more } 
**Everything is a vector.** 

Did you notice that columns in a data frame
are vectors? This means that everything that can be done with a vector can be
done with a data frame column! In fact, all 'atomic' data types in R (numeric,
characters, logical) are *always* stored as vectors -- even a single number is
actually a numerical vector of length one! This may seem like an uninteresting
detail, but it's actually a really powerful idea and the source of much of R's
strength! 
:::

:::{ .digression } 

R has a couple of other data structures that you will eventually encounter. The
details are beyond the scope of this intro, but you can read more
[here.](http://adv-r.had.co.nz/). This includes ...

  - **Factor** -- Factors are numerical vectors that can take a fixed set of
    values (levels), whereby each level can be assigned a label. The levels may
    be ordered or unordered. If this reminds you of categorical and ordinal
    variables you are absolutely right -- that's where they are used most often.
    But be warned: While factors can be quite useful they are also a point of
    confusion for many R novices.
  - **List** -- Lists are similar to vectors, but they can contain a mix of data
    types and data structures, including other lists.
  - [**Tibble**](https://tibble.tidyverse.org/) -- A 'modern' implementation of
    data frames, provided by the Tidyverse.
  - **Object** -- Objects can be used to organize functionality more
    effectively. For instance, the output of a statistical model may be an
    object that has various attributes and associated functions (methods). R
    even has three types of objects: R6, S3, and S4.
  - **Matrix** -- Not to be confused with data frames. If you are implementing
    statistical methods yourself, you will likely want to make use of R's
    powerful matrix algebra functions. Otherwise, you likely won't often have to
    work with matrices directly.
  - [**Environment**](https://adv-r.hadley.nz/environments.html) -- Environments
    are a collection of R objects. They are implemented as a hashmap, which
    makes them a good alternative to named vectors for key-value stores if you
    care about speed. Understanding environments is central to a deep
    understanding of R, but as a beginner it's nothing you need to worry about.
:::
