# A language made for data analysis

::: {.goals}
-   The core data types are numerical, logical, and character
-   To assign data to variables use `<-`
-   Write comments with `#`
-   The most important data structures are vectors and data frames
-   Vectors store a multitude of the same data type in a particular order
-   Data frames are tables with data organized in rows and columns
:::

R is a powerful and complex programming language that supports many different programming
paradigms. However, to get started with data analysis, its enough to understand R's basic
data types --- numeric, logical, and character --- and two data structures, vectors and
data frames. These are made for data manipulation and analysis, and along with the many
convenience functions that R provides, they allow you to do data analysis without worrying
much about things like loops and conditionals, that are fundamental in other programming
languages. As you become more familiar with R, you can add those to your repertoire, too,
and utilize R's full power and flexibility.

## Basic Data Types

The [**data type**](https://en.wikipedia.org/wiki/Data_type) of a piece of data determines
what operations you can perform with it. The bread and butter of data types in R are
*numeric*, *logical*, and *character*.

### Numeric {.unnumbered}

Numeric data can be integer, decimal[^made-for-data-1], or complex, but usually the
difference does not matter. Sometimes, R might display a number in scientific notation,
e.g., `1e+23`. To the uninitiated, these can look cryptic and a bit scary. But don't
worry! It's just a very large or very small number that is written in a more compact form
and if you want you can [change this
behavior.](https://statisticsglobe.com/disable-exponential-scientific-notation-in-r)

[^made-for-data-1]: Decimal numbers are also called floating point numbers or "floats" for
    short. Note that R uses periods `.` to designate the decimal places, not commas `,` as
    you might be used to, depending on where you are from.

```{r, eval=FALSE}
23 + 0.45 + 1e2 + 6i
```

### Logical {.unnumbered}

Logical data (also known as Boolean data) can take only two values, `TRUE` or `FALSE` (or
their shorthands, `T` and `F`). These are often used to represent binary categorical
variables, as function arguments, or as intermediate steps when selecting and filtering
data.

```{r, eval=FALSE}
(TRUE || FALSE) != (F == T)
```

### Character {.unnumbered}

Also known as *strings*. Almost anything you type becomes a character string if you
surround it with `""` or `''`.

```{r, eval=FALSE}
"Hello World!"
```

## Assignment

Often, you want to store some data for later use. For this, you can assign it to a
**variable**[^made-for-data-2]. This is done with the **assignment operator** `<-`. Use
the keyboard shortcut <kbd>Alt/Option</kbd>+<kbd>-</kbd> to save yourself some hand
gymnastics. Once assigned, you can use a variable's name to represent its content in other
operations. Typing a variable name by itself will print the variable's content.

[^made-for-data-2]: In official R terminology, you are assigning values to **objects**,
    but people who have experience with other programming languages sometimes find the
    term object confusing. Therefore, I will use the word *variable* in this book.

```{r}
a <- 1
b <- 9
c <- a + b

c
```

Note how newly created variables appear in the **Environment panel** in RStudio. When you
are working with R, you are working in a **session***.* The session provides an
**environment** in which data is stored, and the variable name allows you to retrieve the
data again. Until you explicitly delete the variable or end the session, the data will be
around for you to use it.

There are some rules for variables names: They must start with a letter and contain only
the letters of the English alphabet, numbers, underscores `_`, hyphens `-`, and dots
`.`.[^made-for-data-3] For long variable names, [it is
recommended](https://style.tidyverse.org/syntax.html#object-names) to use **snake case**,
which looks like this: `my_firssst_variable` ðŸ. Also note that R is case-sensitive.

[^made-for-data-3]: Unlike in other programming languages, dots have no special function
    or meaning in R and can be used in variable names.

::: {.exercise}
What would be the result of running the code below? Think about it first and then run it
to check your answer.

```{r eval=FALSE}
a <- 10
a + b    # What will be the result of a + b?
c        # What has happened to c?
C        # Why is this not working?
```
:::

::: {.more}
Many other programming languages use `=` to assign values to variables. You can, for the
most part, also use `=` for assignment in R --- just be aware that it will let everyone
now that [you are not a true believer](https://en.wikipedia.org/wiki/Shibboleth).

You can also use `->` to assign from left to right. This can it come in handy when you are
messing around in the console, but it may not be a good idea to use it too much in your
scrips, because it can make your code harder to read:

```{r eval=FALSE}
"Never look back!" -> x
```

Finally, there's `<<-`, which allows assignment to variables in an enclosing environment.
This can be useful if you need to update *global variables* from within a function.
However, this is far beyond the scope of this intro, and chances are you will never
encounter a `<<-` in the wild.

You can read more about R's assignment operators
[here](https://colinfay.me/r-assignment/).
:::

## Comments

As you can see in the example above, you can use `#` to leave **comments** in your script.
Everything on the right side of the `#` will be ignored by R. You can quickly "comment
out" multiple lines by going to `Code â–¸ Comment/Uncomment Lines` or pressing
<kbd>Ctrl</kbd>/<kbd>Cmd</kbd> + <kbd>Shift</kbd> + <kbd>C</kbd>. Learn to write good
commentsand write comments gratuitously -- your collaborators and future self will thank
you for it! Sarah Drasner's presentation [*The Art of Code
Comments*](https://www.youtube.com/watch?v=yhF7OmuIILc)is a very entertaining introduction
to good comment-writing.

## Vectors

In computer programming, [**data
structures**](https://en.wikipedia.org/wiki/Data_structure) are a way to organize and
process particular types of data efficiently. When working with data in R, you will
encounter two important data structures: *vectors* and *data frames*. Vectors are made for
storing values of the same data type (numerical, character, logical, ...) in a particular
order.

To create a vector, you can use the "combine" function, `c()`. There are also the `:` and
`seq()` functions, that can help you create a vector that contains a sequence of numbers:

```{r, results="hold"}
c(1,2,3,4)      # create a vector manually
0:4             # a vector of all integer numbers from 0 to 4
4:0             # it also works backwards
seq(0, 2, 0.5)  # numbers 1.0 to 2.0 in 0.5 increments
```

R is really good at working with vectors! See how easy it is to multiply all numbers by a
scalar, get the mean, or sum all elements --- all without using a loop or a temporary
variable:

```{r, results = "hold"}
x <- c(1,2,3,4) # Create vector and assign to variable `x` 
x * 10
mean(x) 
sum(x)
```

You can access specific positions of a vector using square brackets. In R, indices start
at 1, that is, index 1 will return the first element of the vector.

```{r}
x[1]
```

You can also use negative indices to exclude elements. For example, index `-1` will return
everything but the first element of a vector.

```{r}
x[-1]
```

Use `:` within brackets to access a range of indices.

```{r}
x[2:3]
```

You can also use a vector of indices to index into another vector in order to select
specific elements

```{r}
x[c(1,3)]   
```

You can also use the bracket syntax to replace individual elements:

```{r}
x[2:3] <- 0   # replace 2nd and 3rd element with 0
x[2:3] <- c(7,8)   # replace 2nd and 3rd element with 7 and 8, respectively
```

Values in a vector can be named. This comes in handy, for example, if you need a simple
[key-value store:](https://en.wikipedia.org/wiki/Key%E2%80%93value_database)

```{r}
fav_food <- c(
    Garfield = "Lasagna", 
    Popeye = "Spinach",
    Obelix = "Roast Wild Boar"
)

fav_food["Popeye"]
```

::: {.more}
**Implicit Coercion**

Vectors can only contain a *single* data type. If you are trying to mix data types, R will
quietly (without throwing an error) *coerce* (that is, convert) all data types to a data
type that can represent all elements. This coercion follows a certain hierarchy:

-   logical becomes numerical (`FALSE = 0` and `TRUE = 1`)
-   numerical becomes character

The coercion only goes in one direction. For example, character strings that happen to
contain numbers will nevertheless be treated as characters. Consider these examples:

```{r}
c(FALSE, TRUE, 2, 3)   # logical becomes numerical
c(1, 2, "3")           # numerical becomes character
```

This also happens if you update a vector after it has already been created.

```{r}
x <- 1:3 
x[1] <- TRUE   # this logical will become numerical
x 
```

Vectors cannot contain other vectors, so nested vectors automatically get unnested, and as
they get unnested, the elements will be coerced to a common data type:

```{r, results='hold'}
c(1, c(2,3,4)) 
c(c(TRUE, 0), c(TRUE, "0")) # logical and numerical become character
```

This behavior can sometimes be used for good, but in general it's best to avoid it. Make
coercion explicit. For example, if you want to convert a character representation of a
number to numeric, do so with the `as.numeric()` function. For more complex applications
consider static typing with <https://github.com/moodymudskipper/typed>. If you really need
to store a sequence of mixed data types, you will have to use another data structure
called *list*.
:::

## Data Frames

Usually, your research data will be more complex than just a single sequence of values.
For this, R has data frames. Data frames are two-dimensional like a spreadsheet -- with
*rows* usually containing different observations and *columns* usually containing
different variables.

Rows and columns of a data frame can be named, and while row names have fallen out of
fashion, it is good practice to give the columns of your data frames names that are easy
to remember and easy to type.

You can make your own data frame by hand, by putting together vectors like this:

```{r}
cool_numbers <- data.frame(
    "number" = c(1,13,23,42),
    "isPrime" = c(FALSE, TRUE, TRUE, FALSE),
    "oneTwo" = c(1, 2)
)

cool_numbers
```

Note how the vector `c(1,2)` in the `oneTwo` column will get *recycled* to match the
length of the other two columns. This works, as long as the total length is a multiple of
the shorter vector's length.

A quick way to access a single column of a data frame is with the `$` operator.

```{r}
cool_numbers$number
```

Did you notice that columns in a data frame are vectors? This means that everything that
can be done with a vector can be done with a data frame column!

::: {.more}
In fact, all "atomic" data types in R (numeric, characters, logical) are *always* stored
as vectors -- even a single number is actually a numerical vector of length one! This is a
really powerful idea and the source of much of R's strength!
:::

::: {.exercise}
Make a new data frame `my_cool_numbers` with a column `a` that contains all numbers from 1
to 100 and column `b` that contains your favorite number 100 times.

::: {.answer}
``` {.r}
my_cool_numbers <- data.frame(
    "a" = 1:100,
    "b" = 42
)
```
:::

Since data frame columns are vectors, we can use the same operations on them. calculate
the standard deviation of all numbers in column `a` in our `my_cool_numbers` data frame!
Use the RStudio help to find the right function for it.

::: {.answer}
``` {.r}
sd(my_cool_numbers$a)
```
:::
:::

As for vectors, R comes out of the box with many more commands to create and modify data
frames, select columns, and filter rows. You will eventually need to get familiar with
these. However, in this book we will skip these in favor of the more streamlined and
beginner-friendly functions from the *Tidyverse* -- more about that later.

## Other Data Structures

R has a couple of other data structures that you will eventually encounter. I don't want
to dwell on them for too long, because they will not play a role in the rest this book. If
you are interested you can read more [here](http://adv-r.had.co.nz/). Additional R data
structures include ...

-   [**Tibble**](https://tibble.tidyverse.org/) -- You may see the word *Tibble* pop up
    occasionally in the examples in this book. A Tibble is a data structure is a modern
    implementation of data frames. Tibbles are backwards-compatible with data frames and
    for the purpose of this book you can think of them as interchangeable.

-   **Factor** -- Factors are numerical vectors that can take a fixed set of values
    (levels), whereby each level can be assigned a label. The levels may be ordered or
    unordered. If this reminds you of categorical and ordinal variables you are absolutely
    right -- that's where they are used most often. While factors can be quite useful,
    they are also a point of confusion for many R novices.

-   **List** -- Lists are similar to vectors, but they can contain a mix of data types and
    data structures, including other lists.

-   **Object** -- Objects can be used to organize functionality. For instance, a
    statistical model may be an object that has various attributes and associated
    functions. R even has three types of objects: R6, S3, and S4.

-   **Matrix** -- Not to be confused with data frames. If you are implementing statistical
    methods yourself, you may want to make use of R's matrix algebra functions. Otherwise,
    you likely won't often have to work with matrices directly.

-   [**Environment**](https://adv-r.hadley.nz/environments.html) -- Do you remember when I
    said that your R session provides an environment in which data is stored? It turns out
    that environments are yet another data structure in R. An environment is a data
    structure that contains other named R data structures. Understanding environments is
    central to a deep understanding of R, but as a beginner it's nothing you need to worry
    about.
