# Expandability

::: {.goals}
-   R packages are collections of R code written by others
-   The *Tidyverse* is a collection of packages that share a common user-centered design
    philosophy
-   How to install and load an R package?
-   Loading data from a text file with `read_csv()`
-   Saving a data frame to the disk with `write_csv()`
-   Getting a high-level overview of a data frame with `head()`, `view()` and `{skimr}`
:::

R is popular among data analysts, data scientists, researchers, and statisticians. As a
result, it gives you access to a vast amount of amazing packages written by domain
experts, statisticians and programmers from around the world. An **R package** is a
collection of R code that has been bundled up so that it can be easily used by others. The
full power of R unfolds when you tap into this vast ecosystem of packages.

R has an official peer-reviewed **package repository** called
[**CRAN**](https://cran.r-project.org/)as well as domain-specific repositories
[**BioConductor**](https://www.bioconductor.org/), and
[**Neuroconductor**](https://neuroconductor.org/). Many more packages are available on
[Github](https://github.com/topics/r?l=r).

There are tens of thousands of packages to make your life easier! There's an R package for
almost everything! See for yourself at [rdrr.io](https://rdrr.io/) or
[Metacran](https://www.r-pkg.org/). And if there's ever anything missing you can build it
yourself -- together with others across the globe. [It's fun and you will learn a
lot!](https://www.rstudio.com/resources/rstudioglobal-2021/make-a-package-make-some-friends/)
The R Packages book provides a deep dive into writing and publishing R packages.

::: {.more}
By the way: R can also integrate with [Python](https://rstudio.github.io/reticulate/),
[C](http://www.rcpp.org/), and other languages and interface with other software, such as
[MPLUS](https://mclust-org.github.io/mclust/), so you are not limited to what R has to
offer.
:::

## Introducing the Tidyverse

[**The Tidyverse**](https://www.tidyverse.org/) [@tidyverse2019] is a set of R packages
that share a common design philosophy. It aims to provide user-friendly functions for
common data-tasks and tries to iron out some of the idiosyncrasies of **Base R,** that is,
the way R works out of the box. The functions in the Tidyverse have a consistent interface
and easy-to-remember names. The **Tidy Data** that gives the Tidyverse its name has
tabular shape, with

-   one observation per row
-   one variable per column
-   a single-level header
-   variable names that are easy to type and easy to remember

The core idea is this: Once the data is in *tidy* format, all the Tidyverse functions can
be used seamlessly. A special operator, the pipe `%>%`, glues all the functions together,
so that the steps of the analysis can be read from top to bottom, like a recipe (see
below).

The Tidyverse is a relatively recent addition to the R package ecosystem but the design
philosophy has struck a chord with the R developer community and many packages are now
following the same principles. As a result, the overall data analysis workflow in R has
become more streamlined and integrated.

Nevertheless, some people prefer *base R* or other alternatives to the Tidyverse, such as
the [`{data.table}`](https://rdatatable.gitlab.io/data.table/) package. Moreover, many
older packages will never change to a "tidy" design. This leads to a bit of a schism in
the R package ecosystem. Eventually, you will have to learn about these alternatives as
well. However, you can get a long way by just focusing on the Tidyverse.

## Installing packages

For the examples in this book we will use a table with information about climbing
expeditions in the Himalaya that was provided by [Tidy
Tuesday](https://github.com/rfordatascience/tidytuesday/blob/master/data/2020/2020-09-22/readme.md).

Often you will read data from external sources and the first step is to turn it into
something that R can work with, usually a data frame. In this case, the data is provided
as a *CSV* file. CSV stands for comma separated values and is a popular plain text format
for sharing tabular data across the internet.

To read the CSV file and turn it into an R data frame we will use the `read_csv()`
function. To use this function, we first have to install and load the `{tidyverse}`
package.[^expandable-1]

[^expandable-1]: Usually, **packages** are collections of functions. The `{tidyverse}`
    package [@R-tidyverse] is a bit different, as it is actually a collection of other
    packages --- a *meta-package*. When you load it, it loads the most popular packages
    from theTidyverse, such as `{ggplot2}`, `{readr}`, and `{dplyr}`. For example, the
    `read_csv()` function is from the `{readr}` package.

You can install the package from R's central package repository CRAN. The
`install.packages()` function looks for a package of a given name on CRAN, and, if it
finds it, copies it to a special library directory on your computer.

The `library()` function loads a package that you have installed, so that all of its
functions become available for you to use in your current R session. The next time you
start RStudio, you will not have to reinstall the package (it is now on your computer),
but you will have to load it again.

```{r eval=FALSE}
install.packages("tidyverse")
library(tidyverse)
```

::: {.more}
**library vs. package**

The distinction between *library* and *package* can be confusing. It might help to imagine
that you are

-   using `install.packages()`to *add a package to your system's library of packages*
-   using `library()` to *check out a package from your library*, so that you can use it
    in your current R session.

Note also that the quotation marks around the package name in `install.packages()` are
required, because this is essentially a search term that will be used to find the package
on CRAN. The quotation marks in `library()` are optional, because the package exists as an
object in your library, from where you can call it by its name, just like any other
variable.
:::

## Loading data

Now that we have the `{tidyverse}` package installed and loaded, we can use the
`read_csv()` function to load the data straight from the web into an R data frame in our R
session.

```{r eval=FALSE}
peaks <- read_csv("https://raw.githubusercontent.com/teebusch/r-introduction/master/data/peaks.csv")
```

```{r, echo=FALSE}
peaks <- read_csv("data/peaks.csv")  # not shown
```

## Inspecting the data

When we load the data, R shows us the names of the columns in the data set and the type of
each column. We can also have a look at the first few rows with the `head()` function.

```{r}
head(peaks)
```

You can get a more traditional spreadsheet-like view of the complete data frame by
clicking its name in the Environment panel or running the command `view(peaks)`.

To better understand our data, let's get some summary statistics. The `{skimr}` package
[@R-skimr] provides a really nice function for this.

::: {.exercise}
Install and load the `{skimr}` package from CRAN.

::: {.answer}
```{r eval = FALSE}
install.packages("skimr")
library(skimr)
```
:::
:::

Now we can "skim" our data:

```{r}
skim(peaks)
```

This summary tells us quite a lot about our data! For example, we can see that some peaks
have not been successfully climbed until very recently, or not at all. There also appears
to be an outlier in the `first_ascent_year` column that we'll have to take care of -- a
peak that was successfully climbed in the year 201.

::: {.more}
**The :: operator**

We only use one function from the `{skimr}` package, and we only use it this one time. In
a case like this, you might want to use the function without loading the entire package.
To do so, you can put `::` between the package name and the function name. Thus,

```{r eval = FALSE}
skimr::skim(peaks)
```

is equivalent to

```{r eval=FALSE}
library(skimr)
skim(peaks)
```

The `::` is also often used to clarify (either to R or to the person reading the code)
which package a function is coming from. This is particularly useful when there are
functions with the same name in different packages. Another approach to solving this
problem is the [`{conflicted}`](https://conflicted.r-lib.org/) package [@R-conflicted].
:::

## Saving data to disk

We loaded the data from the web straight into R. Thus so far, it only exists in the
current R session, that is, in the computer's working memory. To store a copy of the data
on the hard drive, we can use the `write_csv()` function. It takes the variable and a file
name.

In more complex projects you should put your data files into a subfolder. Here, we will
keep it simple and put everything in the top-level project folder.

```{r, eval=FALSE}
write_csv(peaks, "peaks.csv")
```

::: {.more}
R can connect to many different sorts of data sources, such as
[databases](https://db.rstudio.com/getting-started/connect-to-database/),
[APIs](https://httr.r-lib.org/), and [websites](https://rvest.tidyverse.org/)).\
It can also read and write many different data formats -- including data from software
like SPSS, Stata, and SAS. The `{rio}` package [@R-rio] provides a unified interface for a
whole range of data formats.

Be careful when using Excel to store and share your data! Excel silently converts certain
data. This can [cause
problems](https://www.theverge.com/2020/8/6/21355674/human-genes-rename-microsoft-excel-misreading-dates).
In many cases it may be better to use CSV, JSON,
[Feather](https://blog.rstudio.com/2016/03/29/feather/), or a
[database](https://db.rstudio.com/databases/sqlite/) to store your data.
:::
