---
title: "Ten Reasons to Learn R"
author: "Tobias Busch"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
description: "A short introduction to Contemporary R."
---

```{r include=FALSE}
library(vembedr)
library(tidyverse)
library(skimr)

knitr::opts_chunk$set(
  comment = ""
)
```

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```

# (PART\*) Introduction {.unnumbered}

# Welcome {.unnumbered}

This book is a short hands-on introduction to R. The goal is to provide a quick overview
of things that are great about R so that you can decide whether diving deeper is worth
your time.

I start by briefly introducing the history and purpose of R. Then I present a compelling
reason *not* to learn R and offer some very attractive alternatives. So prepared, we can
then turn to the things that R has to offer: In the second part of the book, I present 10
things that I absolutely love about R. Along with each of the ten reasons, I will
introduce some part of R and will encourage you to get some hands-on experience with R.

We will use modern R that is easy to learn and allows you to focus on your data, rather
than worrying about difficult programming concepts. I will showcase all that R has to
offer, and I hope that this helps you to side whether it is for you.

## Who is this book for? {.unnumbered}

I'm assuming that you are already doing research or data analysis . I assume that you are
already using other tools for that, but you want to learn more about R. Maybe it's
because...

-   You are afraid of missing out and want to know what all the fuzz is about
-   You believe that R might help you to be more efficient
-   You feel that your current workflow isn't reproducible enough
-   There is a statistical method that you would like to use, that is not available in
    your preferred software but it is available in R
-   You unsuccessfully tried to learn R before, but would like to try again

I'm assuming you have little or no programming experience[^index-1] and you are not
particularly interested in computers programming. You don't want to be bored with
technical background knowledge. You are here, because you want to learn the tools that can
help you get things done!

[^index-1]: If you are a seasoned programmer, there are more efficient ways to catch up on
    R. I recommend the books *R for Data Science* and [*Advanced
    R*](https://adv-r.hadley.nz/index.html). Both are available for free online.

## The origins of this book {.unnumbered}

This book started as my notes for a workshop I gave in April 2021 at a research group
meeting at the University of Oslo's Department of Special Needs Education. It was a 90
minutes long hands-on introduction to R. The goal was to provide an overview of R that
could help the workshop participants decide whether they should learn R or not. After the
workshop, I continued to expand on the material and eventually, it turned into a more
comprehensive introduction to R.

This is book is a result of my passion for teaching R and my interest in the paedagogy of
teaching tech. It is heavily inspired by the Carpentries and Greg Wilson's [*Teaching Tech
Together*](https://teachtogether.tech/). However, it tries to be a more compact and
provide a broad overview of R at the expanse of depth of understanding.

## About the author {.unnumbered}

I have worked with R and other languages in a research context for almost 10 years and I
have made some humble contributions to the R ecosystem. I teach R as a certified
instructor for [The
Carpentries](https://www.ub.uio.no/english/writing-publishing/dsc/carpentry-uio/), and I
help maintain the Carpentries lesson material for the lesson *Data Analysis and
Visualization in R for Ecologists*. I really enjoy teaching researchers to code and
introducing them to the many wonders of R.
